---
title: "Tidy time series analysis in R"
date: "27 September 2019"
author: "Rob J Hyndman"
toc: true
output:
  binb::monash:
    colortheme: monashwhite
    fig_width: 8
    fig_height: 5
    highlight: tango
    keep_tex: yes
    includes:
      in_header: preamble.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE
)
library(tidyverse)
library(lubridate)
library(tsibble)
library(feasts)
library(fable)
library(patchwork)
global_economy <- tsibbledata::global_economy %>%
  select(Year, Country, GDP, Imports, Exports, Population)
savepdf <- function(file, width = 16, height = 10) {
  fname <<- paste("figs/", file, ".pdf", sep = "")
  pdf(fname, width = width / 2.54, height = height / 2.54, pointsize = 10, bg = "white")
  par(mgp = c(2.2, 0.45, 0), tcl = -0.4, mar = c(3.3, 3.6, 1.1, 1.1))
}
endpdf <- function() {
  crop::dev.off.crop(fname)
}
gghist <- function(data, mapping, ...) {
  x <- GGally::eval_data_col(data, mapping$x)
  bw <- 0.2 * bw.nrd0(x) + 0.8 * bw.SJ(x)
  p <- ggplot(data, mapping) +
    geom_density(col = NA, fill = "#cc5900", bw = bw)
  return(p)
}
set.seed(20190927)
options(digits = 3, width = 63)
```

```{r tourism, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE,
  dev.args = list(pointsize = 11)
)
  tourism <- tourism %>%
  mutate(
    State = recode(State,
                   "Australian Capital Territory" = "ACT",
                   "New South Wales"="NSW",
                   "Northern Territory" = "NT",
                   "Queensland" = "QLD",
                   "South Australia" = "SA",
                   "Tasmania" = "TAS",
                   "Victoria"="VIC",
                   "Western Australia" = "WA"
    )
  )
holidays <- tourism %>%
  filter(Purpose=="Visiting") %>%
  group_by(State) %>%
  summarise(
    Trips = sum(Trips)
  ) %>%
  select(Quarter, State, Trips)
```

## Tidyverts packages

\begin{textblock}{3.8}(8,0)\begin{alertblock}{}\Large\textbf{tidyverts.org}\end{alertblock}\end{textblock}

\placefig{1}{1.4}{width=4cm}{tsibble.png}
\placefig{5}{1.4}{width=4cm}{tsibbledata.png}
\placefig{3}{4.85}{width=4cm}{feasts.png}
\placefig{7}{4.85}{width=4cm}{fable.png}

# tsibble package

## Time series data

  - Four-yearly Olympic winning times
  - Annual Google profits
  - Quarterly Australian beer production
  - Monthly rainfall
  - Weekly retail sales
  - Daily IBM stock prices
  - Hourly electricity demand
  - 5-minute freeway traffic counts
  - Time-stamped stock transaction data


## `tsibble` objects

\fontsize{10}{11.2}\sf

```{r, echo = TRUE}
global_economy
```

\only<2->{\begin{textblock}{.75}(2.15,3.7)
\begin{alertblock}{}\fontsize{10}{10}\sf Index\phantom{dg}\end{alertblock}
\end{textblock}}
\only<3->{\begin{textblock}{1.6}(3.28,3.7)
\begin{alertblock}{}\fontsize{10}{10}\sf Key\phantom{dg}\end{alertblock}
\end{textblock}}
\only<4>{\begin{textblock}{6.7}(5.5,3.7)
\begin{alertblock}{}\fontsize{10}{10}\sf Measured variables\phantom{dg}\end{alertblock}
\end{textblock}}

## `tsibble` objects

\fontsize{10}{11.3}\sf

```{r, echo = TRUE}
tourism
```

\only<2->{\begin{textblock}{1.1}(2.1,3.7)
\begin{alertblock}{}\fontsize{10}{10}\sf Index\phantom{dg}\end{alertblock}
\end{textblock}}
\only<3->{\begin{textblock}{3.9}(3.65,3.7)
\begin{alertblock}{}\fontsize{10}{10}\sf Keys\phantom{dg}\end{alertblock}
\end{textblock}}
\only<4-5>{\begin{textblock}{1.5}(7.95,3.7)
\begin{alertblock}{}\fontsize{10}{10}\sf Measure\phantom{dg}\end{alertblock}
\end{textblock}}

\only<5>{\begin{textblock}{3}(9,5)
\begin{block}{}\fontsize{10}{10}\sf Domestic visitor nights in thousands by state/region and purpose.\phantom{dg}\end{block}
\end{textblock}}

## `tsibble` objects

* A `tsibble` allows storage and manipulation of multiple time series in R.

* It contains:

  + An index: time information about the observation
  + Measured variable(s): numbers of interest
  + Key variable(s): optional unique identifiers for each series

* It works with tidyverse functions.

# feasts package


## feasts package

\begin{textblock}{3.8}(8,0)\begin{alertblock}{}\Large\textbf{tidyverts.org}\end{alertblock}\end{textblock}

\placefig{1}{1.4}{width=4cm}{tsibble.png}
\placefig{5}{1.4}{width=4cm}{tsibbledata.png}
\placefig{3}{4.85}{width=4cm}{feasts.png}
\placefig{7}{4.85}{width=4cm}{fable.png}


## Holidays by state

\fontsize{9}{10}\sf

```r
holidays <- tourism %>%
  filter(Purpose=="Holiday") %>%
  group_by(State) %>%
  summarise(Trips = sum(Trips))
```

```{r, echo=FALSE}
holidays
```


## Time plots

\fontsize{9}{10}\sf

```{r, echo = TRUE}
holidays %>% autoplot(Trips)
```

## Season plots
\fontsize{9}{10}\sf

```{r graphics1, fig.width=4, fig.height=5}
holidays %>% gg_season(Trips)
```

## Graphics
\fontsize{9}{10}\sf

```{r graphics2}
holidays %>% gg_subseries(Trips)
```

## Graphics
\fontsize{9}{10}\sf

```{r graphics4, fig.width=4, fig.height=5}
holidays %>% ACF(difference(Trips, 4)) %>% autoplot()
```

## Features

\placefig{9.1}{.5}{width=3.6cm}{tukey}
\begin{textblock}{3}(9.7,5.4)\small\textit{John W Tukey}\end{textblock}
\begin{textblock}{8}(0.7,1.2)
\begin{alertblock}{Cognostics}
Computer-produced diagnostics\\ (Tukey and Tukey, 1985).
\end{alertblock}
\end{textblock}\pause
\vspace*{2.5cm}

\alert{Examples for time series}

  * lag correlation
  * size and direction of trend
  * strength of seasonality
  * timing of peak seasonality
  * spectral entropy

\vspace*{10cm}

## Extracting single time series
\fontsize{11}{12}\sf

```{r extract, echo=TRUE}
snowy <- tourism %>%
  filter(
    Region=="Snowy Mountains",
    Purpose=="Holiday"
  )
snowy
```

## Decompositions
\fontsize{9}{10}\sf

```{r dcmp-plot, echo=TRUE}
snowy %>% STL(Trips ~ season(window = "periodic")) %>%
  autoplot()
```

## Candidate features

\begin{block}{STL decomposition}
\centerline{$Y_t = S_t + T_t + R_t$}
\end{block}\pause\fontsize{14}{16}\sf\vspace*{-0.2cm}

* Strength of seasonality: $\max\left(0,1 - \frac{\Var(R_t)}{\Var(Y_t-T_t)}\right)$
* Strength of trend:  $\max\left(0,1 - \frac{\Var(R_t)}{\Var(Y_t-S_t)}\right)$
* Timing of seasonal peaks/troughs
* Linearity/curvature of trend
* Spikiness of remainder
* Autocorrelations of remainder ($R_1,\dots,R_T$)


## Feature extraction and statistics

\fontsize{9}{10}\sf

```{r features, echo=TRUE}
tourism %>% features(Trips, feature_set(tags="stl"))
```

## Feature extraction and statistics
\fontsize{8}{9}\sf

```{r features-plot, fig.height=4.6, echo=TRUE}
tourism %>% features(Trips, feature_set(tags="stl")) %>%
  ggplot(aes(x=trend_strength, y=seasonal_strength_year, col=Purpose)) +
    geom_point() + facet_wrap(vars(State))
```

\only<2->{\begin{textblock}{4.7}(7.8,7.3)
\begin{alertblock}{}\fontsize{10}{10}\sf
\begin{itemize}\tightlist
\item Holidays more seasonal than other travel.
\item WA has strongest trends.
\end{itemize}
\end{alertblock}\end{textblock}}

## Feature extraction and statistics
\fontsize{9}{9}\sf

Find the most seasonal time series:

```{r extreme, echo=TRUE}
most_seasonal <- tourism %>%
  features(Trips, feature_set(tags="stl")) %>%
  filter(seasonal_strength_year == max(seasonal_strength_year))
```

\pause\vspace*{-0.3cm}

```{r extreme2, fig.height=3., echo=TRUE}
tourism %>%
  right_join(most_seasonal, by = c("State","Region","Purpose")) %>%
  ggplot(aes(x = Quarter, y = Trips)) + geom_line() +
  facet_grid(vars(State,Region,Purpose))
```

```{r pca, echo=FALSE}
# Save pdf figures
savepdf <- function(file, width=16, height=10)
{
  fname <<- paste("figs/",file,".pdf",sep="")
  pdf(fname, width=width/2.54, height=height/2.54, pointsize=10)
  par(mgp=c(2.2,0.45,0), tcl=-0.4, mar=c(3.3,3.6,1.1,1.1))
}
endpdf <- function()
{
  crop::dev.off.crop(fname)
}
# Compute features
tourism_features <- tourism %>%
  features(Trips, feature_set(pkgs="feasts"))
# Compute PCs
pcs <- tourism_features %>% select(-State, -Region, -Purpose) %>%
  prcomp(scale=TRUE) %>% augment(tourism_features)
# Save some PC plots
savepdf("pca1", 18, 10)
pcs %>% ggplot(aes(x=.fittedPC1, y=.fittedPC2)) +
  geom_point() + theme(aspect.ratio=1)
endpdf()
savepdf("pca2", 18, 10)
pcs %>% ggplot(aes(x=.fittedPC1, y=.fittedPC2, col=State)) +
  geom_point() + theme(aspect.ratio=1)
endpdf()
savepdf("pca3", 18, 10)
pcs %>% ggplot(aes(x=.fittedPC1, y=.fittedPC2, col=Purpose)) +
  geom_point() + theme(aspect.ratio=1)
endpdf()
# Find outliers
outliers <- pcs %>%
  filter(.fittedPC1 == max(.fittedPC1) |
         (.fittedPC1 > 10 & .fittedPC2 > 2.5))
savepdf("pca4", 18, 10)
pcs %>% ggplot(aes(x=.fittedPC1, y=.fittedPC2, col=Purpose)) +
  geom_point() + theme(aspect.ratio=1) +
  geom_point(data=outliers, aes(x=.fittedPC1, y=.fittedPC2), col="black", shape=1, size=3)
endpdf()
```

## Feature extraction and statistics
\fontsize{9}{9}\sf

```{r tourismfeatures, echo=TRUE}
tourism_features <- tourism %>%
  features(Trips, feature_set(pkgs="feasts"))
```

```{r echo=FALSE}
tourism_features
```

\begin{textblock}{2.5}(9.6,1.3)
\begin{alertblock}{}\fontsize{10}{12}\sf
All features from the feasts package
\end{alertblock}
\end{textblock}

## Feature extraction and statistics
\fontsize{9}{9}\sf

```{r pcatable, echo=TRUE}
pcs <- tourism_features %>% select(-State, -Region, -Purpose) %>%
  prcomp(scale=TRUE) %>% augment(tourism_features)
```

```{r echo=FALSE}
pcs
```

\begin{textblock}{2.5}(9.6,3)
\begin{alertblock}{}\fontsize{10}{12}\sf
Principal components based on all features from the feasts package
\end{alertblock}
\end{textblock}

## Feature extraction and statistics
\fontsize{9}{9}\sf

\begin{textblock}{3.3}(.4,3)
\begin{alertblock}{}\fontsize{10}{12}\sf
Principal components based on all features from the feasts package
\end{alertblock}
\end{textblock}

```r
pcs %>% ggplot(aes(x=.fittedPC1, y=.fittedPC2)) +
  geom_point() + theme(aspect.ratio=1)
```

\placefig{4}{2.6}{height=6.4cm, width=12cm}{pca1}
\vspace*{10cm}

## Feature extraction and statistics
\fontsize{9}{9}\sf

\begin{textblock}{3.3}(.4,3)
\begin{alertblock}{}\fontsize{10}{12}\sf
Principal components based on all features from the feasts package
\end{alertblock}
\end{textblock}

```r
pcs %>% ggplot(aes(x=.fittedPC1, y=.fittedPC2, col=State)) +
  geom_point() + theme(aspect.ratio=1)
```

\placefig{4}{2.6}{height=6.4cm, width=12cm}{pca2}
\vspace*{10cm}

## Feature extraction and statistics
\fontsize{9}{9}\sf

\begin{textblock}{3.3}(.4,3)
\begin{alertblock}{}\fontsize{10}{12}\sf
Principal components based on all features from the feasts package
\end{alertblock}
\end{textblock}

```r
pcs %>% ggplot(aes(x=.fittedPC1, y=.fittedPC2, col=Purpose)) +
  geom_point() + theme(aspect.ratio=1)
```

\only<1>{\placefig{4}{2.6}{height=6.4cm, width=12cm}{pca3}}
\only<2>{\placefig{4}{2.6}{height=6.4cm, width=12cm}{pca4}}
\vspace*{10cm}

## Feature extraction and statistics
\fontsize{9}{9}\sf

```{r outliers2, fig.height=3.4}
pcs %>%
  filter(.fittedPC1 == max(.fittedPC1)) %>%
  left_join(tourism, by = c("State", "Region", "Purpose")) %>%
  ggplot(aes(x = Quarter, y = Trips)) +
    geom_line() +
    facet_grid(vars(State,Region,Purpose)) +
    ggtitle("Outlying time series in PC space") +
    theme(legend.position = "none")
```

## Feature extraction and statistics
\fontsize{9}{9}\sf

```{r outliers3, fig.height=3.4}
pcs %>%
  filter(.fittedPC1 > 10 & .fittedPC2 > 2.5) %>%
  left_join(tourism, by = c("State", "Region", "Purpose")) %>%
  ggplot(aes(x = Quarter, y = Trips)) +
    geom_line() +
    facet_grid(vars(State,Region,Purpose)) +
    ggtitle("Outlying time series in PC space") +
    theme(legend.position = "none")
```


# fable package

## fable package

\begin{textblock}{3.8}(8,0)\begin{alertblock}{}\Large\textbf{tidyverts.org}\end{alertblock}\end{textblock}

\placefig{1}{1.4}{width=4cm}{tsibble.png}
\placefig{5}{1.4}{width=4cm}{tsibbledata.png}
\placefig{3}{4.85}{width=4cm}{feasts.png}
\placefig{7}{4.85}{width=4cm}{fable.png}


## Model estimation

The `model()` function estimates models for tsibbles.

\fontsize{9}{9}\sf

```{r scaleup2}
tourism %>%
  model(
    snaive = SNAIVE(Trips),
    ets = ETS(Trips),
    arima = ARIMA(Trips)
  )
```

###
A `mable` is a model table, each cell corresponds to a fitted model.


## Producing forecasts
\fontsize{10}{12}\sf

```{r scaleup3}
tourism %>%
  model(
    snaive = SNAIVE(Trips),
    ets = ETS(Trips),
    arima = ARIMA(Trips)
  ) %>%
  forecast(h= "3 years")
```

###
A `fable` is a forecast table with point forecasts and distributions.

## Training and test sets

```{r traintest, fig.height=1, echo=FALSE, cache=TRUE}
train = 1:18
test = 19:24
par(mar=c(0,0,0,0))
plot(0,0,xlim=c(0,26),ylim=c(0,2),xaxt="n",yaxt="n",bty="n",xlab="",ylab="",type="n")
arrows(0,0.5,25,0.5,0.05)
points(train, train*0+0.5, pch=19, col="blue")
points(test,  test*0+0.5,  pch=19, col="red")
text(26,0.5,"time")
text(10,1,"Training data",col="blue")
text(21,1,"Test data",col="red")
```

  * A model which fits the training data well will not necessarily forecast well.
  * Forecast accuracy is based only on the test set.

### Forecast errors

Forecast "error": the difference between an observed value and its forecast.
$$
  e_{T+h} = y_{T+h} - \hat{y}_{T+h|T},
$$
where the training data is given by $\{y_1,\dots,y_T\}$

## Forecast errors

```{r trainall, echo=TRUE}
train <- tourism %>%
  filter(year(Quarter) <= 2014)
fit <- train %>%
  model(
    ets = ETS(Trips),
    arima = ARIMA(Trips),
    snaive = SNAIVE(Trips)
  ) %>%
  mutate(mixed = (ets+arima+snaive)/3)
fc <- fit %>% forecast(h="3 years")
```

## Forecast errors
\fontsize{10}{12}\sf

```{r snowytrain, dependson='trainall', fig.height=4}
fc %>%
  filter(Region=="Snowy Mountains", Purpose=="Holiday") %>%
  autoplot(level=NULL) +
  autolayer(snowy, Trips)
```

## Measures of forecast accuracy

\begin{tabular}{rl}
$y_{T+h}=$ & $(T+h)$th observation, $h=1,\dots,H$ \\
$\pred{y}{T+h}{T}=$ & its forecast based on data up to time $T$. \\
$e_{T+h} =$  & $y_{T+h} - \pred{y}{T+h}{T}$
\end{tabular}

\begin{align*}
\text{MAE} &= \text{mean}(|e_{T+h}|) \\[-0.2cm]
\text{MSE} &= \text{mean}(e_{T+h}^2) \qquad
&&\text{RMSE} &= \sqrt{\text{mean}(e_{T+h}^2)} \\[-0.1cm]
\text{MAPE} &= 100\text{mean}(|e_{T+h}|/ |y_{T+h}|)
\end{align*}\pause

  * MAE, MSE, RMSE are all scale dependent.
  * MAPE is scale independent but is only sensible if $y_t\gg 0$ for all $t$, and $y$ has a natural zero.

## Measures of forecast accuracy

\begin{block}{Mean Absolute Scaled Error}
$$
\text{MASE} = \text{mean}(|e_{T+h}|/Q)
$$
where $Q$ is a stable measure of the scale of the time series $\{y_t\}$.
\end{block}
Proposed by Hyndman and Koehler (IJF, 2006).

For non-seasonal time series,
$$
  Q = (T-1)^{-1}\sum_{t=2}^T |y_t-y_{t-1}|
$$
works well. Then MASE is equivalent to MAE relative to a naïve method.

\vspace*{10cm}

## Measures of forecast accuracy

\begin{block}{Mean Absolute Scaled Error}
$$
\text{MASE} = \text{mean}(|e_{T+h}|/Q)
$$
where $Q$ is a stable measure of the scale of the time series $\{y_t\}$.
\end{block}
Proposed by Hyndman and Koehler (IJF, 2006).

For seasonal time series,
$$
  Q = (T-m)^{-1}\sum_{t=m+1}^T |y_t-y_{t-m}|
$$
works well. Then MASE is equivalent to MAE relative to a seasonal naïve method.

\vspace*{10cm}

## Measures of forecast accuracy

\fontsize{9}{10}\sf

```{r snowy-test-accuracy, dependson='trainall'}
accuracy(fc, tourism)
```


## Measures of forecast accuracy

\fontsize{10}{10}\sf

```{r snowy-test-accuracy2, dependson='trainall'}
accuracy(fc, tourism) %>%
  group_by(.model) %>%
  summarise(
    RMSE = mean(RMSE),
    MAE = mean(MAE),
    MASE = mean(MASE)
  ) %>%
  arrange(RMSE)
```

## Acknowledgements

\begin{block}{}
\fontsize{11}{11}\sf
\centering\begin{tabular}{l@{\hspace*{1cm}}l}
\includegraphics[height=4cm, width=10cm]{mitch2} &
\includegraphics[height=4cm, width=10cm]{earowang} \\
Mitchell O'Hara-Wild &
Earo Wang \\
\end{tabular}
\end{block}\vspace*{-0.15cm}

\begin{alertblock}{}
\centerline{\textbf{tidyverts.org}}
\centerline{\textbf{robjhyndman.com}}
\end{alertblock}


\vspace*{10cm}